---
title: "US Migration Data"
output:
  pdf_document:
    fig_width: 7
    fig_height: 3
    fig_caption: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
library(fanetc)
library(rstan)
library(truncnorm)
library(bayesplot)
library(tidybayes)
library(distortr)
load('./data_setup2.RData')

reg_d <- lapply(reg_d, function(dd){
  dd %>% 
  filter(age<=75) %>% 
  mutate(age_gp = cut(age, breaks = seq(0,75,5))) %>% 
  group_by(age_gp) %>% 
  mutate(age = median(age)) %>% 
  group_by(newpuma, year, age_gp,age) %>% 
  summarise(migrant = sum(migrant), 
            totalpop = sum(totalpop)) %>% 
  mutate(rate = migrant/totalpop) %>% 
  ungroup()
})

reg_d$`6`$statefip = 6
reg_d$`36`$statefip = 36
reg_d$`48`$statefip = 48

reg_d$`6`$state = 'California'
reg_d$`36`$state = 'New York'
reg_d$`48`$state = 'Texas'

aa <- paste0(reg_d$`36` %>% group_by(age) %>% slice(1)%$% age_gp, collapse=", ")

# reg_d$`36`
# reg_d$`36` <- reg_d$`36` %>% 
#   left_join(obs_counties, by = c("newpuma","statefip"))

 reg_d$`36` %>% mutate(dd = totalpop-migrant) %>% arrange(-rate) 
 
 set.seed(36)
 # sim_pop <- 500
 sim_pop <- 1000
  reg_d$`36` <- reg_d$`36` %>% 
   rowwise() %>% 
   mutate(sim_dat = rpois(1, rate*sim_pop)) %>% 
   ungroup() %>% 
   mutate(sim_pop = sim_pop,
          sim_rate = sim_dat/sim_pop)
```

# Couple of descriptive plots

```{r, echo = FALSE, warning=FALSE}
reg_d$`36` %>% 
  ggplot(aes(year, rate, group = age_gp, col = age_gp))+ 
  geom_line(show.legend = FALSE)+ 
  facet_wrap(~newpuma, scales = "free")+ 
  theme_bw()
```

```{r, echo = FALSE, warning=FALSE}
reg_d$`36` %>% 
  group_by(newpuma, age_gp) %>% 
  mutate(rate = rate - mean(rate)) %>% 
  ggplot(aes(year, rate, group = newpuma, col = newpuma))+ 
  geom_line(show.legend = FALSE)+ 
  facet_wrap(~age_gp)+ 
  theme_bw()
```

```{r, echo = TRUE, warning=FALSE}

lm1 <- glm(migrant ~ age_gp + factor(newpuma) + offset(log(totalpop)), data = reg_d$`36`, family = poisson(link="log"))
fitted_links = log(predict(lm1, type="response")/reg_d$`36`$totalpop)
rate_resid = reg_d$`36`$rate - predict(lm1, type="response")/reg_d$`36`$totalpop

reg_d$`36` %>% 
  group_by(newpuma, age_gp) %>% 
  mutate(rate = rate - mean(rate)) %>% 
  ggplot(aes(year, rate, group = newpuma, col = newpuma))+ 
  geom_line(show.legend = FALSE)+ 
  facet_wrap(~age_gp)+ 
  theme_bw()


reg_d$`36` %>% 
  mutate(rate_resid = rate_resid) %>% 
  group_by(newpuma, age_gp) %>% 
  # mutate(rate = rate - mean(rate)) %>% 
  ggplot(aes(year, rate_resid, group = newpuma, col = newpuma))+ 
  geom_line(show.legend = FALSE)+ 
  facet_wrap(~age_gp)+ 
  theme_bw()

reg_d$`36` %>% 
  mutate(rate_resid = rate_resid) %>% 
  group_by(newpuma, age_gp) %>% 
  # mutate(rate = rate - mean(rate)) %>% 
  ggplot(aes(age, rate_resid, group = newpuma, col = newpuma))+ 
  geom_line(show.legend = FALSE)+ 
  facet_wrap(~year)+ 
  theme_bw()
```


# Models for New York for years 2013-2019

Define $y_{a,r,t}$ to be the number of persons of age $a$ migrating in to region $r$ at year $t$. Define $P_{a,r,t}$ to be the population aged $a$ in region $r$ at year $t$. We assume migrant counts are Poisson distributed as follows: 
$$y_{a,r,t} \sim \text{Poisson}(\mu_{a,r,t}P_{a,r,t})$$

## Model 1.2

We model the rates on the log scale with form: 
$$\log \mu_{r,a,t} = \alpha+ \tau_{a} + \gamma_r + \delta_{r,a,t}.$$


The full model is: 
\begin{align*}
y_{a,r,t}&\sim \text{Poisson}(\mu_{a,r,t}P_{a,r,t})\\
\log \mu_{a,r,t} &= \alpha + \tau_a + \gamma_r  +  \delta_{a,r,t}\\
\delta_{a,r,t}&= \delta^1_{a,r} + \delta^2_{r,t} + \delta^3_{a,r,t}\\
\delta^1_{a,r} &\sim \mathcal{N}(\delta^1_{a-1,r},\sigma_{\delta^1_r}^2)\\
\delta^2_{r,t} &\sim \mathcal{N}(\delta^2_{r,t-1},\sigma_{\delta^2_r}^2)\\
\dleta^3_{a,r,t} & \sim \mathcal{N}(0,\sigma_3^2)\\
&\sum_{r}\delta^1_{a,r} = 0\\
&\sum \delta^3 = 0\\
\tau_a &\sim \mathcal{N}(0,1)\\
\gamma_r &\sim \mathcal{N}(0,1)\\
\sigma_{\delta^1_r} &\sim \mathcal{N}^+(0,1)\\
\sigma_{\delta^2_r} &\sim \mathcal{N}^+(0,1)\\
\sigma_3 &\sim \mathcal{N}^+(0,1)
\end{align*}

```{r, echo = FALSE, warning = FALSE, message= FALSE}


tmp <- reg_d$`36` %>% 
  filter(year <=2019)

tmp <- tmp %>% 
  mutate(age_gp_num = as.numeric(as.factor(age_gp)), 
         region = as.numeric(as.factor(newpuma)))

# tmp <- tmp %>%
# filter(region <= 6)

ages <- unique(tmp$age) 
I <- 10 # between-knot length -->
res <- distortr::GetSplines(ages, I = I) # a function from distortr, to get splines of constant shape 
B <- res$B.ik

A = length(unique(tmp$age_gp))
R = length(unique(tmp$newpuma))
N = length(tmp$migrant)
T = length(unique(tmp$year))
age_unique = tmp$age_gp %>% unique()
region_unique = tmp$newpuma %>% unique()
pred_df = expand.grid(age_gp = age_unique, 
                      newpuma = region_unique) %>% 
  mutate(year = 2020)

stan_data <- list(A = A,
                  R = R,
                  N = N,
                  T = T,
                  K= ncol(B),
                  age = model.matrix(~tmp$age_gp)[,-1],
                  region = model.matrix(~ as.factor(tmp$newpuma))[,-1],
                  time = model.matrix(~ as.factor(tmp$year))[,-1],
                  age_region = model.matrix(~tmp$age_gp*as.factor(tmp$newpuma))[,-c(1:(R+A-1))],
                  pop = log(tmp$sim_pop),
                  y = tmp$sim_dat,
                  # pop = log(tmp$totalpop),
                  # y = tmp$migrant,
                  age_vec = tmp$age_gp %>% as.numeric(),
                  region_vec = as.factor(tmp$newpuma) %>% as.numeric(),
                  time_vec = as.factor(tmp$year) %>% as.numeric(),
                  pred_age = model.matrix(~pred_df$age_gp)[,-1],
                  pred_region = model.matrix(~ as.factor(pred_df$newpuma))[,-1],
                  pred_age_region = model.matrix(~pred_df$age_gp*as.factor(pred_df$newpuma))[,-c(1:(R+A-1))],
                  pred_age_vec = as.numeric(pred_df$age_gp),
                  pred_region_vec = as.numeric(as.factor(pred_df$newpuma)),
                  B=B)
```

```{r eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
# Random walk on the deviations + random intercepts


modcs <- stan(data = stan_data,
             file = "../Programs/models/final models/finalmodel_spline.stan",
             iter=1000,
             warmup = 250,
             # iter = 300,
             cores=4,
             chains = 4,
             control = list(adapt_delta = 0.95,max_treedepth = 15),
             thin=3,
             # thin = 1,
             seed=1)
model_sum <- summary(modcs)
interaction = FALSE
save(file="../Model results/finalmodel_spline_moreits.RData", list=c("modcs","model_sum", "interaction"))

```


```{r, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
load("../Model results/finalmodel_spline_moreits.RData")
interaction = FALSE
d <- model_sum$summary[,c("mean", "n_eff", "Rhat","sd","se_mean","2.5%","97.5%")]
d %>% as.data.frame() %>% arrange(desc(Rhat)) %>% head(10)
totrace <- rownames(d %>% as.data.frame() %>% arrange(desc(Rhat)))[1]
rstan::traceplot(modcs, pars=totrace,inc_warmup = TRUE)

#extract age-specific rates (tau)
est_tau <- model_sum$summary[,c("mean", "sd", "n_eff", "Rhat")][grep("tau\\[[0-9]*\\]",rownames(d)),1]

#extract region-specific rates (gamma)
est_gamma <- model_sum$summary[,c("mean", "sd", "n_eff", "Rhat")][grep("gamma\\[[0-9]*\\]",rownames(d)),1]

#extract intercept  (alpha)
est_alpha <- model_sum$summary[,c("mean", "sd", "n_eff", "Rhat")][grep("alpha",rownames(d)),1]

#extract log rates
r1 <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("log_rate",rownames(d)),1]
r1_upper <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("log_rate",rownames(d)),3]
r1_lower <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("log_rate",rownames(d)),2]

#extract predictions intervals 
# p1 <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("rate_p",rownames(d)),1]
# p1_upper <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("rate_p",rownames(d)),3]
# p1_lower <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("rate_p",rownames(d)),2]
# p1_se <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("rate_p",rownames(d)),2]

# extract poisson_rng outcomes
yp_upper <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("y_p",rownames(d)),3]
yp_lower <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("y_p",rownames(d)),2]

#extract fixed effects 
f1 <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("fixed_eff",rownames(d)),1][1:nrow(tmp)]

#extract RW on age + fixed effects 
f2 <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("deviation_fixed_eff",rownames(d)),1][1:nrow(tmp)]


#extract predictions intervals year 2020
p2020 <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("pred_rate",rownames(d)),1]
p2020_lower <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("pred_rate",rownames(d)),2]
p2020_upper <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("pred_rate",rownames(d)),3]

#generated
p1.1 <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("generated_rate",rownames(d)),1]
p1.1_upper <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("generated_rate",rownames(d)),3]
p1.1_lower <- model_sum$summary[,c("mean", "2.5%","97.5%")][grep("generated_rate",rownames(d)),2]



est_age_rates <- c(est_alpha + mean(est_gamma), est_tau + est_alpha + mean(est_gamma))
est_age_rates <- exp(est_age_rates)

mig_sched = pred_df %>% dplyr::select(age_gp, newpuma)
mig_sched$fit = model.matrix(~age_gp + as.factor(newpuma), data = mig_sched) %*% (matrix(c(est_alpha, est_tau,est_gamma)))

tmp_results <- tmp

tmp_results$est_log_rates = r1
tmp_results$est_log_rates_upper = r1_upper
tmp_results$est_log_rates_lower = r1_lower
tmp_results$fixed_eff = f1
tmp_results$deviation_fixed_eff = f2
tmp_results$gen_log_rates = p1.1
tmp_results$gen_log_rates_lower = p1.1_lower
tmp_results$gen_log_rates_upper = p1.1_upper


pred_results <- pred_df %>% 
  left_join(reg_d$`36` %>% dplyr::select(newpuma, age_gp, year, rate))

pred_results$pred_log_rates = p2020
pred_results$pred_log_rates_lower = p2020_lower
pred_results$pred_log_rates_upper = p2020_upper


tmp_results <- tmp_results %>% 
  mutate(est_rates = exp(est_log_rates),
         fixed_eff_rates = exp(fixed_eff),
         deviation_fixed_eff = exp(deviation_fixed_eff),
         fitted_rate_lower = exp(est_log_rates_lower),
         fitted_rate_upper = exp(est_log_rates_upper)) %>%
  mutate(log_rate_obs = log(rate))

pred_results <- pred_results %>% 
  mutate(est_rates = exp(pred_log_rates),
         fitted_rate_upper = exp(pred_log_rates_upper),
         fitted_rate_lower = exp(pred_log_rates_lower)
         ) %>%
  mutate(log_rate_obs = log(rate))

tmp_results <- full_join(tmp_results, pred_results)

```

### Estimated standard age schedule

```{r, echo = FALSE, warning=FALSE, message=FALSE}
in_mig_plot <- mig_sched %>% 
  group_by(age_gp) %>% 
  ggplot(aes(age_gp, exp(fit), col = factor(newpuma), group = newpuma)) + 
  geom_line(alpha=0.2, show.legend = FALSE)+ 
  theme_bw() +
  ggtitle("Migration Puma In-Migration Standard Age Schedule, New York") +
  scale_y_continuous(name ="rate",
                     breaks = scales::pretty_breaks(n=10))+
  xlab("age")+ 
  geom_line(data=data.frame(age = tmp$age_gp %>% unique(),
           fit = est_age_rates,
           newpuma = NA), aes(age, fit) ,inherit.aes = TRUE,col = "black")
```

```{r, echo = FALSE, warning=FALSE, message=FALSE}
data.frame(age = tmp$age %>% unique(),
           fit = est_age_rates) %>% 
ggplot(aes(age,fit))+ 
  geom_line()+ 
  theme_bw()
```

